---
title: "Getting Started with rebalancedNoise"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Getting Started with rebalancedNoise}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## The EZS Method

The `rebalancedNoise` package implements the EZS method as described by @Sabolova2025. It applies record-level noise with a dynamic rebalancing algorithm to preserve data quality in non-sensitive cells while ensuring statistical disclosure control (SDC) across complex hierarchies.

## Example with Multiple Variables

The engine can process multiple numerical variables. Initializing them together ensures the underlying SDC problem structure is consistent.

```{r setup}
library(rebalancedNoise)
library(data.table)
library(sdcHierarchies)

# Generate dummy microdata
set.seed(1)
dt <- data.table(
  ID = 1:200,
  REGION = sample(c("North", "South"), 200, replace = TRUE),
  INDUSTRY = sample(c("Tech", "Mfg", "Retail"), 200, replace = TRUE),
  turnover = runif(200, 10, 1000),
  assets = runif(200, 50, 5000),
  direction = sample(c(1, -1), 200, replace = TRUE),
  noise_multiplier = 0.05 
)

# Define hierarchies
dims <- list(
  REGION = hier_create("Total", nodes = c("North", "South")),
  INDUSTRY = hier_create("Total", nodes = c("Tech", "Mfg", "Retail"))
)

# Initialize and perturb
sdc <- rn_setup(
  data = dt, 
  dim_list = dims, 
  num_vars = c("turnover", "assets"), 
  sensitive_params = list(n_threshold = 5)
)

sdc$perturb("turnover")
sdc$perturb("assets")
```

## Flexible Output Formats

The `get_results()` method provides two primary ways to view your data.

### Long Format

Ideal for automated processing or visualization with `ggplot2`. It uses standardized value columns and a `variable` column to distinguish between `turnover` and `assets`.

```{r long}
res_long <- sdc$get_results(format = "long")

# Standardized names: val_orig, val_pert, diff_final_pct
head(res_long[, .(variable, REGION, val_orig, val_pert, is_sens)])
```

### Wide Format (Default)

The wide format pivots the data so that each hierarchical cell occupies a single row. Variable names are used as prefixes for the values.

```{r wide}
res_wide <- sdc$get_results(format = "wide")

# Clean prefixes: [var]_orig, [var]_pert, [var]_is_sens
head(res_wide[, .(REGION, INDUSTRY, turnover_orig, turnover_pert, assets_orig, assets_pert)])

# Identify cells where rebalancing was most effective for turnover
res_wide[turnover_is_sens == FALSE & abs(turnover_diff_final_pct) < 0.01]
```

## Output Variable Description

The following variables are included in the results to help interpret the perturbation quality:

| Variable | Description |
|:---|:---|
| `n_obs` | Number of microdata records contributing to the specific cell. |
| `is_internal` | `TRUE` for base cells (leaf nodes); `FALSE` for hierarchical aggregates. |
| `[var]_orig` | The original aggregated value before perturbation. |
| `[var]_pert` | The final perturbed and rebalanced value. |
| `[var]_is_sens` | `TRUE` if the specific variable was flagged as sensitive in that cell. |
| `[var]_diff_final_pct` | The percentage deviation between original and final values. |

## Evaluation of Results

Before finalizing your tables, use the `$summarize()` method to evaluate the impact of the perturbation. This provides a high-density diagnostic report across three groups: **Overall**, **Non-Sensitive** (Rebalanced), and **Sensitive** (Fixed Noise).

```{r summary}
# Display statistics for the turnover variable
sdc$summarize("turnover")
```

### How to Interpret the Summary

* **MAPE (Initial vs. Final):** In the Non-Sensitive group, the *"Initial" MAPE* reflects the noise before rebalancing. The *"Final" MAPE* should be significantly lower, demonstrating the algorithm's effectiveness.
* **Percentile Comparison:** By comparing the *"Initial"* and *"Final" Relative (%) distributions*, you can visualize the *"shrinking"* effect. A successful run will show the interquartile range (`25%` to `75%`) moving towards `0`.
* **Tail Risks:** Monitor the `1%` and `99%` percentiles. These identify if specific cells in the hierarchy received disproportionately high noise, which can occasionally occur in very small or isolated nodes.
* **Absolute Impact:** The *"Absolute (Units)"* distribution translates these percentages into the original scale of your data, helping you assess the actual impact on table utility.

## Performance and Parallelization

The EZS engine utilizes C++ with OpenMP for computationally intensive tasks. The number of threads (`n_threads`) is resolved during `rn_setup()` and remains fixed for the life of the object.

The engine resolves the thread count using the following priority:

1.  **Function Argument:** `rn_setup(..., n_threads = 4)`
2.  **Global Option:** `options(rn_threads = 4)`
3.  **Environment Variable:** `Sys.setenv(rn_threads = "4")`
4.  **Automatic Fallback:** `parallel::detectCores() - 1` (Minimum of 1)

```{r threads_example}
# Example: Setting a session-wide preference
options(rn_threads = 4)

# Initialize - this object is now locked to 4 threads
sdc <- rn_setup(dt, dims, "turnover")
```

## Summary of Logic

* **Hierarchy Integrity:** Aggregations across the hierarchy remain additive.
* **Automatic Cleanup:** Internal SDC identifiers are removed, leaving only the dimensions and results.
* **Calculated Deviations:** Percentage deviations are rounded to 3 decimal places for immediate quality checks.

## References
