---
title: "Getting Started with rebalancedNoise"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Getting Started with rebalancedNoise}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## The EZS Method

The `rebalancedNoise` package implements the EZS method as described by
@Sabolova2025. The method adds record-level noise to numeric survey variables.
Every record carries a random direction (either 1 or -1) and a noise
multiplier. Cells identified as sensitive keep their fixed direction, which
guarantees protection. In non-sensitive cells the algorithm adjusts the
directions so that the noise cancels out within table hierarchies. The result
is disclosure control with low distortion and additive consistency across all
aggregation levels.

## Two Ways to Work

The package offers two interchangeable interfaces. Both produce identical
results and can be mixed freely.

1. The **R6 engine**. `rn_setup()` creates one object that keeps all state.
   You work with methods like `$rebalance()` and `$perturb()`. Tables are
   stored under names, results are cached, and one rebalancing serves any
   number of publication tables.
2. The **functional interface**. Plain functions take a typed object and
   return a new one. You chain them with the pipe operator `|>`:
   `rn_init()` returns an `rn_initialized` object, `rn_rebalance()` returns
   an `rn_rebalanced` object, and `rn_perturb()` returns an `rn_perturbed`
   object. There is no hidden state, so each pipeline is self-contained.

The table below maps the corresponding entry points.

| Task | R6 engine | Functional interface |
|:---|:---|:---|
| Initialize | `rn_setup()` | `rn_init()` |
| Rebalance | `sdc$rebalance()` | `rn_rebalance()` |
| Perturb | `sdc$perturb()` | `rn_perturb()` |
| Retrieve a table | `sdc$get_results()` | `rn_format()` |
| Evaluate a table | `sdc$summarize()` | `rn_summarize()` |
| Microdata with directions | `sdc$get_microdata()` | `as.data.table(state)` |
| Reuse in another session | `sdc$export()` | `saveRDS()` and `rn_init()` |
| Start over after rebalancing | `sdc$reset()` | rebalance again, no reset needed |
| Overview of stored tables | `sdc$list_tables()` | not needed, you hold the objects |

The R6 engine suits interactive work where you create many tables from the
same rebalanced microdata and want caching and naming support. The functional
interface suits scripts and reproducible pipelines because each step is an
explicit, testable function call. The vignette
`vignette("workflows", package = "rebalancedNoise")` covers all details of
both styles.

## Setup

The following data is used throughout this vignette. It contains two
dimensions (country and industry) and three numeric variables (turnover,
workers, assets). Country probabilities are unequal on purpose so that some
cells hold fewer than five records and are flagged as sensitive.

```{r setup}
library(rebalancedNoise)
library(data.table)
library(sdcHierarchies)

set.seed(1)
N <- 300
countries <- c("AT", "DE", "NL", "SE", "FR", "IT")
industries <- c("Tech", "Mfg", "Retail", "Services")

country_probs <- c(0.25, 0.25, 0.20, 0.15, 0.10, 0.05)

dt <- data.table(
  country = sample(countries, N, replace = TRUE, prob = country_probs),
  industry = sample(industries, N, replace = TRUE),
  turnover = runif(N, 1000, 10000),
  workers = sample(10:500, N, replace = TRUE),
  assets = runif(N, 10000, 100000),
  direction = sample(c(1, -1), N, replace = TRUE),
  noise_multiplier = sample(seq(0.75, 1.25, by = 0.01), N, replace = TRUE)
)

# Detailed hierarchy for rebalancing (country x industry)
dims_detailed <- list(
  country = hier_create("Total", nodes = countries),
  industry = hier_create("Total", nodes = industries)
)

# Coarse hierarchies for publication tables
dims_country_only <- list(
  country = hier_create("Total", nodes = countries)
)

dims_industry_only <- list(
  industry = hier_create("Total", nodes = industries)
)

head(dt)

# Cell sizes, the smallest cells will be sensitive
dt[, .N, by = .(country, industry)][order(N)]
```

Every input record must provide the two EZS columns `direction` (1 or -1) and
`noise_multiplier` (0 or greater). Set `Sys.setenv(SDC_LOG_LEVEL = "OFF")` if
you do not want the engine to print status messages.

## The Same Workflow Twice

The workflow has three steps: initialize the data, rebalance once on the
detailed structure, and perturb a publication table at a coarser level. The
two following sections run exactly this workflow, first with the R6 engine
and then with the functional interface.

### Using the R6 engine

```{r quick-start-r6}
# One object holds all state
sdc <- rn_setup(data = dt, sensitive_params = list(n_threshold = 5))

# Rebalance once on the detailed structure
sdc$rebalance(dim_list = dims_detailed, num_var = "turnover")

# Perturb a publication table at country level, stored under a name
sdc$perturb(dim_list = dims_country_only, variables = "turnover", name = "table_a")

# Read the stored table
res_r6 <- sdc$get_results("table_a")
head(res_r6)
```

### Using the functional interface

```{r quick-start-fun}
# Each step returns a typed object, the pipe passes it to the next step
res_fun <- dt |>
  rn_init(sensitive_params = list(n_threshold = 5)) |>
  rn_rebalance(dim_list = dims_detailed, num_var = "turnover") |>
  rn_perturb(dim_list = dims_country_only, variables = "turnover")

# The pipeline result is an rn_perturbed object
res_fun

# Convert it to a data.table in wide format
res_fun_wide <- rn_format(res_fun)
head(res_fun_wide)
```

Both routes compute the same table, and the results are identical:

```{r quick-start-identical}
identical(as.data.frame(res_r6), as.data.frame(res_fun_wide))
```

The intermediate object `rn_perturbed` carries the hierarchy, the SDC
parameters, and a hash, so the pipeline stays reproducible without any
external bookkeeping. Each function checks the class of its input strictly
and rejects wrong types with a hint about the required preceding step. A
plain `data.table`, even one that contains a `direction_rebalanced` column,
is not accepted by `rn_perturb()`.

## Reading the Results

Both interfaces provide the same output formats, and the resulting
`data.table` columns are the same in both cases. Use `sdc$get_results()` or
`rn_format()` with `format = "wide"` (the default) or `format = "long"`.

### Wide Format

Wide format has one row per table cell. Every perturbed variable contributes
a group of columns.

```{r wide-format}
head(res_r6)
```

### Long Format

Long format has one row per cell and variable, with standardized column
names. It is convenient for plotting and further processing.

```{r long-format}
head(sdc$get_results("table_a", format = "long"))
head(rn_format(res_fun, format = "long"))
```

### Column Reference

| Format | Column | Content |
|:---|:---|:---|
| both | `[dim]` | dimension value, one column per dimension |
| both | `n_obs` | number of records contributing to the cell |
| both | `is_internal` | `TRUE` for base cells, `FALSE` for aggregates |
| wide | `[var]` | unperturbed original value |
| wide | `[var]_init` | perturbation with the original random directions |
| wide | `[var]_pert` | final perturbed value after rebalancing |
| wide | `is_sens_[var]` | `TRUE` if the cell is sensitive for that variable |
| long | `variable` | name of the variable of this row |
| long | `val_orig`, `val_pert_init`, `val_pert` | the three values of `[var]`, `[var]_init`, and `[var]_pert` |
| long | `is_sensitive` | `TRUE` if the cell is sensitive for this variable |
| long | `diff_init_pct`, `diff_final_pct` | percentage deviation of the initial and final values from the original |

The comparison between the three versions of each variable is what makes the
method auditable. `[var]` is the truth, `[var]_init` shows what plain
perturbation would have produced, and `[var]_pert` shows the final published
value.

You can restrict the output to selected variables with the `variables`
argument, which both `sdc$get_results()` and `rn_format()` accept. A variable
that is not part of the table causes an error that lists the available ones.

## Combining Both Interfaces

The two interfaces are interchangeable at every point in the workflow. An R6
object can hand its state to the functional interface, and an R6 result table
can be evaluated with the functional tools.

```{r combine}
# Feed the R6 engine state into the functional pipeline
state <- sdc$get_state()  # rn_rebalanced
t_b <- rn_perturb(state, dim_list = dims_industry_only, variables = "workers")
head(rn_format(t_b))

# Evaluate a stored R6 table with the functional summary
rn_summarize(sdc$get_table("table_a"))
```

The reverse direction works as well. An object created with `rn_init()` can
be passed to `rebalancedNoise$new()` to continue with R6 methods.

## Next Steps

The vignette `vignette("workflows", package = "rebalancedNoise")` describes
the remaining features in both styles: multiple variables and multiple
tables from one rebalancing, rebalancing on linear combinations, table naming
and caching in the R6 engine, rounding, evaluation with `summarize()`,
export and import, and parallelization.

## References
