---
title: "Perturbation Workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Perturbation Workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

This vignette covers the detailed features of `rebalancedNoise` after the
basic workflow. Every feature is shown in both interfaces: the stateful R6
engine (`rn_setup()` plus `$` methods) and the functional interface
(`rn_init()` plus piped function calls). The basics of both styles are
explained in `vignette("getting-started", package = "rebalancedNoise")`.

## Setup

The same simulated data is used throughout this vignette. Country
probabilities are unequal so that some cells hold fewer than five records.

```{r setup}
library(rebalancedNoise)
library(data.table)
library(sdcHierarchies)

set.seed(1)
N <- 300
countries <- c("AT", "DE", "NL", "SE", "FR", "IT")
industries <- c("Tech", "Mfg", "Retail", "Services")

country_probs <- c(0.25, 0.25, 0.20, 0.15, 0.10, 0.05)

dt <- data.table(
  country = sample(countries, N, replace = TRUE, prob = country_probs),
  industry = sample(industries, N, replace = TRUE),
  turnover = runif(N, 1000, 10000),
  workers = sample(10:500, N, replace = TRUE),
  assets = runif(N, 10000, 100000),
  direction = sample(c(1, -1), N, replace = TRUE),
  noise_multiplier = sample(seq(0.75, 1.25, by = 0.01), N, replace = TRUE)
)

dims_detailed <- list(
  country = hier_create("Total", nodes = countries),
  industry = hier_create("Total", nodes = industries)
)

dims_country_only <- list(
  country = hier_create("Total", nodes = countries)
)

dims_industry_only <- list(
  industry = hier_create("Total", nodes = industries)
)
```

## One Rebalancing, Many Tables

Sensitive cell detection runs once during rebalancing and is reused by every
following perturbation. Rebalancing is the expensive step, so the choice of
the detailed hierarchy deserves attention. Publication tables can then use
any coarser hierarchy and any numeric variable without further rebalancing.

The chunks below set up the two core objects that the rest of this vignette
builds on. Both objects rebalance `turnover` on the detailed structure
(country x industry) and create one country-level table.

```{r core-r6}
sdc <- rn_setup(data = dt, sensitive_params = list(n_threshold = 5))
sdc$rebalance(dim_list = dims_detailed, num_var = "turnover")
sdc$perturb(dim_list = dims_country_only, variables = "turnover", name = "table_a")
```

```{r core-fun}
state_fun <- dt |>
  rn_init(sensitive_params = list(n_threshold = 5)) |>
  rn_rebalance(dim_list = dims_detailed, num_var = "turnover")

t_a <- rn_perturb(state_fun, dim_list = dims_country_only, variables = "turnover")
```

Additional tables reuse the same rebalanced state. In the R6 engine every
table is stored under a name inside the object:

```{r many-tables-r6}
# A second table, different hierarchy, two variables in one call
sdc$perturb(
  dim_list = dims_industry_only,
  variables = c("turnover", "assets"),
  name = "table_b"
)
res_b <- sdc$get_results("table_b")
head(res_b[, list(
  industry, turnover, turnover_init, turnover_pert,
  assets, assets_init, assets_pert
)])

# All stored tables at once, returned as a named list
names(sdc$get_results())
```

In the functional interface you keep the objects yourself, for example in a
named list. The `rn_rebalanced` object `state_fun` can be perturbed as often
as needed:

```{r many-tables-fun}
t_b <- rn_perturb(
  state_fun,
  dim_list = dims_industry_only,
  variables = c("turnover", "assets")
)

tables <- list(table_a = rn_format(t_a), table_b = rn_format(t_b))
names(tables)
```

There is no automatic caching in the functional interface. A repeated call to
`rn_perturb()` recomputes the table. In the R6 engine a repeated call with
the same name and the same parameters is recognized and skipped.

## Sensitive Cell Detection

The results carry the sensitivity flags that the rebalancing step derived
from `sensitive_params`. Cells with few contributing records are flagged, and
their noise direction stays fixed while the algorithm optimizes all other
directions.

```{r sensitive-cells}
res_a <- sdc$get_results("table_a")
res_a[, list(country, n_obs, is_internal, is_sens_turnover)][order(n_obs)]
```

The `n_obs` column counts the records contributing to each cell. The
`is_internal` column marks base cells, and `is_sens_turnover` reports the
sensitivity decision for that variable. In long format the same information
appears in the standardized `is_sensitive` column.

## Table Naming and Caching in the R6 Engine

Every R6 result table is identified by its name and by a hash of the
hierarchy, the SDC parameters, and the rounding flag. The hash enables two
protections.

Repeating an identical call recomputes nothing:

```{r hash-cache}
sdc$list_tables()

# Same name, same hash: the calculation is skipped
sdc$perturb(dim_list = dims_country_only, variables = "turnover", name = "table_a")
```

Reusing a name for a different table structure raises an error instead of
silently overwriting a stored result:

```{r hash-collision}
# Different hierarchy under an existing name: error
try(sdc$perturb(dim_list = dims_industry_only, variables = "turnover", name = "table_a"))

# Different structures need different names
sdc$perturb(dim_list = dims_industry_only, variables = "turnover", name = "table_c")
sdc$list_tables()
```

The functional interface has no table registry. You store each `rn_perturbed`
object under a name of your own choosing, so collisions cannot occur.

## Rounding to Whole Numbers

Some publications require integer values. The `round` argument, available in
`sdc$perturb()` and `rn_perturb()`, rounds the perturbed values on the
microdata level with R's `round()` before the tables are computed. Because
rounding happens before tabulation, every published value is a whole number.

```{r round-r6}
sdc$perturb(
  dim_list = dims_country_only, variables = "turnover",
  name = "table_rounded", round = TRUE
)
res_rounded <- sdc$get_results("table_rounded")
head(res_rounded[, list(country, turnover, turnover_pert)])

all(res_rounded$turnover_pert == round(res_rounded$turnover_pert))
```

```{r round-fun}
t_rounded <- rn_perturb(
  state_fun,
  dim_list = dims_country_only, variables = "turnover", round = TRUE
)
res_rounded_fun <- rn_format(t_rounded)
all(res_rounded_fun$turnover_pert == round(res_rounded_fun$turnover_pert))
```

Four points to keep in mind:

- Only the perturbed columns (`[var]_init` and `[var]_pert`) are rounded.
  The original values in `[var]` stay unchanged.
- Rounding adds up to 0.5 per record, so cell totals can shift a little.
- R's `round()` rounds halves toward the nearest even number, so 2.5 becomes
  2. Exact halves rarely occur in perturbed data.
- The round flag is part of the R6 table hash. Reusing a table name with a
  different `round` value raises a collision error.

## Rebalancing on Linear Combinations

A single variable is not always the right basis for direction optimization.
If turnover, workers, and assets matter equally, rebalancing on only one of
them protects the others less. In that case, build a weighted linear
combination of standardized variables and use it as the rebalancing target.
The algorithm then protects all variables together.

Missing values need attention here. A missing variable should neither add to
nor subtract from the combined score, and dropping records with missing
values wastes information. The recommended solution is to replace missing
values with 0 before scaling. After centering, 0 equals the mean, so missing
values contribute neutrally.

```{r linear-comb-data}
# Work on a copy so that dt stays complete
dt2 <- copy(dt)
dt2[sample(1:N, 20), assets := NA]
sum(is.na(dt2$assets))

# Standardize turnover and workers, replace NA with 0 before scaling assets
dt2[, turnover_std := scale(turnover)]
dt2[, workers_std := scale(workers)]
dt2[, assets_std := scale(ifelse(is.na(assets), 0, assets))]

# Weighted combination: 30 percent turnover, 30 percent workers, 40 percent assets
dt2[, rebal_combined := 0.3 * turnover_std + 0.3 * workers_std + 0.4 * assets_std]
sum(is.na(dt2$rebal_combined))
```

The combined variable is used only in the rebalancing step. The perturbation
still operates on the original variables.

```{r linear-comb-r6}
sdc_lc <- rn_setup(data = dt2, sensitive_params = list(n_threshold = 5))
sdc_lc$rebalance(dim_list = dims_detailed, num_var = "rebal_combined")
sdc_lc$perturb(
  dim_list = dims_country_only,
  variables = c("turnover", "workers", "assets"),
  name = "multi_var_table"
)
res_lc <- sdc_lc$get_results("multi_var_table")
head(res_lc[, list(country, turnover, turnover_pert, workers, workers_pert)])
```

```{r linear-comb-fun}
res_lc_fun <- dt2 |>
  rn_init(sensitive_params = list(n_threshold = 5)) |>
  rn_rebalance(dim_list = dims_detailed, num_var = "rebal_combined") |>
  rn_perturb(
    dim_list = dims_country_only,
    variables = c("turnover", "workers", "assets")
  )
head(rn_format(res_lc_fun)[, list(country, turnover, turnover_pert)])
```

## Evaluating Results

Before releasing tables, inspect the perturbation impact. The R6 method
`$summarize()` and the function `rn_summarize()` report the same statistics.

```{r evaluate}
sdc$summarize(table = "table_a", target_vars = "turnover")
```

```{r evaluate-fun}
rn_summarize(t_a)
```

How to read the report:

- **MAPE** is the mean absolute percentage error between original and
  perturbed values. MAPE (Initial) refers to the noise with the original
  random directions, before rebalancing. MAPE (Final) refers to the published
  values. A lower final value shows the effect of the rebalancing.
- **Noise Reduction** states by how many percent the final MAPE is below the
  initial MAPE.
- **Percentiles** list nine levels of the deviation distribution: 0, 1, 5,
  25, 50, 75, 95, 99, and 100 percent. The first two rows are relative
  deviations in percent, initial and final. The third row gives absolute
  deviations in the original unit of the variable. Check the 1 and 99 percent
  levels for extreme deviations in single cells.
- **Sensitivity Rate** states which share of cells was flagged as sensitive.
  These cells received fixed noise for disclosure protection.

Both functions return the statistics invisibly as a named list, so you can
store and process them. With a single `target_vars` entry the list contains
the elements `overall` and `meta`, with one element per variable otherwise:

```{r summarize-structure}
result_a <- sdc$summarize(table = "table_a", target_vars = "turnover")
names(result_a)
names(result_a$overall)
names(result_a$meta)

result_b <- sdc$summarize(table = "table_b", target_vars = c("turnover", "assets"))
names(result_b)
```

## Export and Import

Rebalancing large datasets takes time. The rebalanced state can be stored and
loaded again so that later sessions only pay for the perturbations.

### Using the R6 engine

`$export()` writes the rebalanced microdata, the SDC parameters, the
rebalancing status, and, optionally, the cached result tables. `rn_setup()`
recognizes an export object or the path to an RDS file and restores the full
state.

```{r export-r6}
export_file <- file.path(tempdir(), "rebalanced_data.rds")
sdc$export(file = export_file)

# In a later session: load and continue without rebalancing again
sdc2 <- rn_setup(export_file)
sdc2$list_tables()

# New tables from the imported state, no recomputation of the directions
sdc2$perturb(dim_list = dims_industry_only, variables = "workers", name = "table_d")
sdc2$list_tables()
```

Pass `include_results = FALSE` to export only the rebalanced microdata, which
produces a smaller file. On import, `rn_setup()` validates the object and
rejects files with missing columns or invalid direction values.

### Using the functional interface

An `rn_rebalanced` object contains everything that is needed to perturb
further tables, so a plain `saveRDS()` is enough:

```{r export-fun}
state_file <- file.path(tempdir(), "rebalanced_state.rds")
saveRDS(state_fun, state_file)

state_re <- readRDS(state_file)
t_re <- rn_perturb(state_re, dim_list = dims_industry_only, variables = "workers")
head(rn_format(t_re))
```

The R6 export object can also be fed directly into the functional interface,
which makes the two persistence formats interchangeable:

```{r export-mixed}
rn_init(sdc$export(include_results = FALSE))
```

## Utility Methods

### Microdata with Directions

`sdc$get_microdata()` returns the microdata including `direction`,
`direction_rebalanced`, and `noise_multiplier`. Set
`include_record_id = TRUE` to add the internal record identifier that the
rebalancing uses for tracking. The functional counterpart is
`as.data.table()` applied to an `rn_rebalanced` object.

```{r microdata}
head(sdc$get_microdata())
head(sdc$get_microdata(include_record_id = TRUE))

# Functional equivalent
head(as.data.table(state_fun))
```

### Listing Stored Tables

`sdc$list_tables()` shows the name, the dimensions, and the variables of
every table stored in the R6 object.

```{r list-tables}
sdc$list_tables()
```

### Starting Over with reset()

Rebalancing can run once per R6 object and once per imported export state. A
second call raises an error. `$reset()` returns the object to its initialized
state: it removes the rebalancing artifacts (`direction_rebalanced`, `strID`,
and `is_sens_*` columns), restores the original row order, and clears all
cached tables. Optionally it accepts new SDC parameters.

```{r reset}
sdc_rst <- rn_setup(dt, sensitive_params = list(n_threshold = 5))
sdc_rst$rebalance(dim_list = dims_detailed, num_var = "turnover")
sdc_rst$perturb(dim_list = dims_detailed, variables = "turnover", name = "t1")

# A second rebalance() on the same object is rejected
try(sdc_rst$rebalance(dim_list = dims_country_only, num_var = "turnover"))

# Reset, optionally with new parameters, then start again
sdc_rst$reset(sensitive_params = list(n_threshold = 3))
sdc_rst$rebalance(dim_list = dims_country_only, num_var = "turnover")
sdc_rst$perturb(dim_list = dims_country_only, variables = "turnover", name = "t1")
sdc_rst$list_tables()
```

After `$reset()` the object behaves like a freshly initialized one. The
functional interface needs no equivalent: `rn_rebalance()` is stateless and
can be applied to an `rn_rebalanced` object again at any time.

## Performance and Parallelization

The sensitivity check during rebalancing is the most time-consuming step and
runs in C++ with OpenMP. It is used in `$rebalance()` and `rn_rebalance()`, in
`$perturb()` with `strID`, and in `rn_perturb()` when the sensitivity flags of
a result table are computed. The perturbation step itself runs single-threaded.

The number of threads is resolved once in `rn_setup()` or `rn_init()` and then
stays fixed for the object or state. Besides the `n_threads` argument there,
the value can also be set globally with `options(rn_threads = 4)` or via the
environment variable `rn_threads`, for example with `Sys.setenv(rn_threads = "4")`.
The argument takes precedence over the option, and the option takes precedence
over the environment variable. Without any of these settings, the default is
the number of detected cores minus one, with a minimum of 1. This setting
applies to both interfaces.

## Summary

Every feature of `rebalancedNoise` is reachable through both interfaces, and
the interfaces can be combined at any stage of a workflow. The R6 engine adds
naming, caching, and state management for interactive table production. The
functional interface provides explicit, composable steps for scripts and
reproducible pipelines. Choose by working style, not by capability.
